\chapter{Conclusions}
\label{chap:FutureWork}

In this project, an algorithm to solve the timetabling problem, was proposed. It was the main objective to use the problem described in the \gls{itc2007}, and so the instances presented in this competition were used. The developed algorithm utilizes an hybridization of heuristics (\gls{gc}) and meta-heuristics (\gls{sa} and \gls{hc}), to solve the problem.\\
\\
With the use of this hybrid approach of \gls{gc}, \gls{sa}, and \gls{hc} we've reached near the results of the five finalists of the \gls{itc2007} on most of the datasets. In some datasets, we were able to be right before M\"{u}ller's results, the winner of the competition.\\
\\
Our results were compared against the top five contestants of the competition, as well as with state-of-the-art approaches. These latter approaches were able to beat M\"{u}ller's results on most of the datasets, but M\"{u}ller's approach can still get very good results on all the datasets. M\"{u}ller's results contrast with some approaches that can beat his results on some datasets but get much worse results on others.\\
\\
We can conclude that the \gls{sa} works as predicted, getting much better results on most of the datasets when executed for much more time. Running the current approach for 12 hours, yielded better results, as compared to M\"{u}ller's and some recent approaches.\\
\section{Future Work}
There are different approaches to be tested with the implemented approach in order to improve the results. Some new approaches may include the use of more heuristics with the \gls{sa} and \gls{hc}, such as \gls{gd}. One improvement that must be implemented in order to get better results is editing the implemented \gls{sa} and make it able to move more than one examination at each time, to allow examinations with \textit{exam\_coincidence} hard constraint to be moved. It is expected that after implementing this improvement, the number of rejected moves might get lower, leading to produce better results.\\
\\
An extensive study should take place, in order to try and find feasible solutions for the dataset 4. To reach this goal, different approaches should be implemented and tested using the \gls{gc}.